---
title: "Make tree mutation burden figure"
output: html_notebook
---

# Metadata
* **Last modified:** `r format(Sys.time(), "%Y%m%d")`

## Purpose
* Plot mutation burden on full tissue tree (log-scaled weights) and germ layer tree (raw weights)
* Create a table of edge weights (full tissue and germ layer trees)
* Plot predominant mutation type on full tree and germ layer trees
* Plot var type distribution for each parent-sibling group and test for differences among spatial and temportal dimensions


# Setup environment
```{r}
library(XNomial)
library(MASS)
```


```{r}
source("../tree_utils.R")
source("../utils.R")
source("../fig.tree_burden/mk_tree_mutation_burden.functions.R")
```
## Globals
```{r}
local = 2
today = "final" # get_todays_date()
fig_dir = "output_files/"
save_file = FALSE
#save_file = TRUE

vartype_palette = create_vartype_palette(local)
vartype_palette_df = load_vartype_colors(local)

compressed_var = names(generate_compressed_var()$compressed2uncompressed)

var_types_lookup = generate_compressed_var()$uncompressed2compressed
```

# Full tree
```{r}
tree_name = "full.no_gtex_blacklist.no_sex.no_cell_lines"
consensus_g_fn = "../fig.tree_burden/input_files/consensus_g.power_adjusted_weights.exact_match.full.no_gtex_blacklist.no_sex.no_cell_lines.20200714.rds"
mutation_type_dist_fn = "input_files/var_type_distribution.power_adjusted_weights.exact_match.full.no_gtex_blacklist.no_sex.no_cell_lines.20200714.rds"
edge_indexes_to_exclude = c(9, 46, 58)

tmp = load_data(tree_name, consensus_g_fn, local, edge_indexes_to_exclude)

tissue_g = tmp$tissue_g
consensus_g = tmp$consensus_g
leaves = tmp$leaves
non_leaf_edge_indexes = tmp$non_leaf_edge_indexes
distance_from_root = tmp$distance_from_root
mutation_vector_str2edge = tmp$mutation_vector_str2edge
```

## Mutation burden
### Log-scaled weights
```{r, fig.width=20, fig.height=20}
fn = file.path(fig_dir, sprintf("mutation_burden.no_labels.log_weights.%s.%s.pdf", tree_name, today))

if (save_file) {
    pdf(fn, width=20, height=20)
}
consensus_g_log_weights = plot_g(consensus_g, type="mutation_prob", category="numeric", save_file=FALSE, set_na_colors = TRUE, set_widths=FALSE, diverging_palette=FALSE, set_edge_labels=FALSE, na_color="grey", edge_label_color="black", set_vertex_labels=FALSE, use_log_trans = TRUE, pseudo_count = 0)
full_tree_consensus_g_log_weights = consensus_g_log_weights
if (save_file) {
    dev.off()
    write_new_file_message(fn)
}
```

```{r}
# Get range of mutation weights
unknown_weights = E(consensus_g)$weight_unknown
edge_mutation_weights = E(consensus_g)$mutation_prob
woi = edge_mutation_weights[! unknown_weights]
writeLines(sprintf("min weight = %.1E", min(woi)))
writeLines(sprintf("max weight = %.1E", max(woi)))

writeLines(sprintf("top 2 edges have %.2f of the weight", sum(sort(woi, decreasing = TRUE)[1:2])))
writeLines(sprintf("top 3 edges have %.2f of the weight", sum(sort(woi, decreasing = TRUE)[1:3])))
```

```{r}
top_woi = sort(edge_mutation_weights, decreasing = TRUE, index.return=TRUE)$ix[1:3]
E(consensus_g)[top_woi]
```

Make a color bar
```{r, fig.height=7, fig.width=2}
fn = file.path(fig_dir, sprintf("mutation_burden.no_labels.log.weights.%s.color_bar.%s.pdf", tree_name, today))
mk_color_bar(consensus_g_log_weights, fn, log_trans=TRUE, percent=TRUE, save_file=save_file)
```

### Create a table of edge weights
```{r}
unknown_weights = E(consensus_g)$weight_unknown
edge_mutation_weights = E(consensus_g)$mutation_prob

edge_weight_df = as.data.frame(ends(consensus_g, E(consensus_g), names=TRUE))
colnames(edge_weight_df) = c("edge_head", "edge_tail")
edge_weight_df$edge_mutation_weight = edge_mutation_weights
edge_weight_df$edge_mutation_weight[unknown_weights] = NA
```

```{r}
# # Run 1X on 20220629
# output_fn = file.path(fig_dir, sprintf("mutation_burden.weights.%s.%s.txt", tree_name, today))
# write.table(edge_weight_df, file=output_fn, row.names=FALSE, quote=FALSE, sep="\t")
# write_new_file_message(output_fn)
```

## Mutation type
```{r}
tmp = load_mutation_type_data(mutation_type_dist_fn, mutation_vector_str2edge, edge_indexes_to_exclude)

mutation_type_dist = tmp$mutation_type_dist
mutation_type_dist_melt = tmp$mutation_type_dist_melt
mutation_type_dist_filtered = tmp$mutation_type_dist_filtered
```


### Test for differences in mutation type distributions across edges
```{r}
compare_mut_type_between_edges_df = compare_mutation_type_distributions_in_tree(mutation_type_dist, mutation_vector_str2edge, "mc_multinomial", leaves$leaf_names, compressed_var)
```


```{r, fig.width=10, fig.height=4}
plot_mutation_type_distribution_comparisons(mutation_type_dist_melt, distance_from_root, tree_name, today, fig_dir, save_file=save_file)
```

### Test for predomiant mutation type within an edge
```{r}
predominant_edge_df = test_if_predominant_type_exists(mutation_type_dist, compressed_var)
```

```{r, fig.width=20, fig.height=20}
fn = file.path(fig_dir, sprintf("mutation_type_distribution.predominant_type.%s.%s.pdf", tree_name, today))
if (save_file) {
    pdf(fn, width=20, height=20)
}
plot_predominant_mutation_type(predominant_edge_df, consensus_g, vartype_palette_df)

if (save_file) {
    dev.off()
    write_new_file_message(fn)
}
```

# Germ layer tree
```{r}
tree_name = "germ_layers"
consensus_g_fn = "../fig.tree_burden/input_files/consensus_g.power_adjusted_weights.exact_match.germ_layers.20200715.rds"
mutation_type_dist_fn = "input_files/var_type_distribution.power_adjusted_weights.exact_match.germ_layers.20200715.rds"
edge_indexes_to_exclude = c()

tmp = load_data(tree_name, consensus_g_fn, local, edge_indexes_to_exclude)

tissue_g = tmp$tissue_g
consensus_g = tmp$consensus_g
leaves = tmp$leaves
non_leaf_edge_indexes = tmp$non_leaf_edge_indexes
distance_from_root = tmp$distance_from_root
mutation_vector_str2edge = tmp$mutation_vector_str2edge
```

## Mutation burden
### Raw weights
```{r, fig.width=20, fig.height=20}
fn = file.path(fig_dir, sprintf("mutation_burden.no_labels.raw_weights.%s.%s.pdf", tree_name, today))

if (save_file) {
    pdf(fn, width=20, height=20)
}
consensus_g_raw_weights = plot_g(consensus_g, type="mutation_prob", category="numeric", save_file=FALSE, set_na_colors = TRUE, set_widths=FALSE, diverging_palette=FALSE, set_edge_labels=FALSE, na_color="grey", edge_label_color="black", set_vertex_labels=FALSE, use_log_trans = FALSE, pseudo_count = 0)
germ_layer_tree_consensus_g_raw_weights = consensus_g_raw_weights
if (save_file) {
    dev.off()
    write_new_file_message(fn)
}
```

Make a color bar
```{r, fig.height=7, fig.width=2}
fn = file.path(fig_dir, sprintf("mutation_burden.no_labels.raw_weights.%s.color_bar.%s.pdf", tree_name, today))
mk_color_bar(consensus_g_raw_weights, fn, log_trans=FALSE, percent=TRUE, save_file=save_file)
```

### Create a table of edge weights
```{r}
unknown_weights = E(consensus_g)$weight_unknown
edge_mutation_weights = E(consensus_g)$mutation_prob

edge_weight_df = as.data.frame(ends(consensus_g, E(consensus_g), names=TRUE))
colnames(edge_weight_df) = c("edge_head", "edge_tail")
edge_weight_df$edge_mutation_weight = edge_mutation_weights
edge_weight_df$edge_mutation_weight[unknown_weights] = NA
```

```{r}
# # Run 1X on 20220629
# output_fn = file.path(fig_dir, sprintf("mutation_burden.weights.%s.%s.txt", tree_name, today))
# write.table(edge_weight_df, file=output_fn, row.names=FALSE, quote=FALSE, sep="\t")
# write_new_file_message(output_fn)
```

## Mutation type
```{r}
tmp = load_mutation_type_data(mutation_type_dist_fn, mutation_vector_str2edge, edge_indexes_to_exclude)

mutation_type_dist = tmp$mutation_type_dist
mutation_type_dist_melt = tmp$mutation_type_dist_melt
mutation_type_dist_filtered = tmp$mutation_type_dist_filtered
```

### Test for differences in mutation type distributions across edges
```{r}
compare_mut_type_between_edges_df = compare_mutation_type_distributions_in_tree(mutation_type_dist, mutation_vector_str2edge, "mc_multinomial", leaves$leaf_names, compressed_var)
```


```{r, fig.width=3, fig.height=4}
plot_mutation_type_distribution_comparisons(mutation_type_dist_melt, distance_from_root, tree_name, today, fig_dir, save_file=save_file, width=3, height=4)
```


### Test for predomiant mutation type within an edge
```{r}
predominant_edge_df = test_if_predominant_type_exists(mutation_type_dist, compressed_var)
```

```{r, fig.width=20, fig.height=20}
fn = file.path(fig_dir, sprintf("mutation_type_distribution.predominant_type.%s.%s.pdf", tree_name, today))
if (save_file) {
    pdf(fn, width=20, height=20)
}
plot_predominant_mutation_type(predominant_edge_df, consensus_g, vartype_palette_df)

if (save_file) {
    dev.off()
    write_new_file_message(fn)
}
```