---
title: "Model mutation burden"
output: html_notebook
---

# Metadata
* **Last modified:** `r format(Sys.time(), "%Y%m%d")`

## Purpose
* Fit a regression model for sample tissue burden
* Plot regression coefficients for significant covariates
* Plot F statistics for model
* Estimate the percent of donor variance explained

## Note
* In this analysis, self-reported ancestry is used for modeling mutation burden. We note that genetic ancestry is an imperfect measurement of genetic diversity and has several limitations (Lewis et al. Science 2022).  Additionally, the variable "RACE" is used for self-reported ancestry metadata.  The variable name was used to maintain consistency with the original metadata files from the GTEx consortium.

## Output files
* `output_files/model.<model>.<date>.rds` = RDS of fitted model using model=mutation count as mutation burden model=mutation VAF sum as mutation burden
* `output_files/<model>.tissue_ancestry_effects.african_american_vs_white.<date>.pdf` = regression coefficients for significant tissue-AA vs. EA interactions for each model
* `output_files/<model>.tissue_ancestry_effects.asian_vs_white.<date>.pdf` = regression coefficients for significant tissue-AS vs. EA interactions for each model
* `output_files/<model>.tissue_ancestry_effects.asian_vs_african_american.<date>2.pdf` = regression coefficients for significant tissue-AS vs. AE interactions for each model
* `output_files/<model>.tissue_sex_effects.<date>.pdf` = regression coefficients for significant tissue-sex interactions for each model
* `output_files/<model>.tissue_age_effects.<date>.pdf` = regression coefficients for significant tissue-age interactions for each model

# Setup environment
```{r}
library("ggplot2")
library("gridExtra")
library("car")
library("grid")
```

```{r}
source("../utils.R")
```

## Globals
```{r}
local = 2
tissue_palette = create_tissue_palette(local)
today = get_todays_date()

input_dir = "input_files"
output_dir = "output_files"
fig_dir = output_dir

save_fig_bool = FALSE
# save_fig_bool = TRUE

metadata = load_master_sample_metadata(local)
metadata = metadata[metadata$is_hypermutated == 0,]

# Calculate mutation burden as the sum as VAF in a sample
# Since burden will be log-transformed; add a pseudocount (smallest non-zero vaf) to all samples
vaf_sum_pseudocount = min(metadata$sample_vaf_sum[metadata$sample_vaf_sum > 0])
metadata$sample_vaf_sum_pseudo = metadata$sample_vaf_sum + vaf_sum_pseudocount

metadata_all_races = metadata
# Use White as reference since it has the largest sample size
metadata_all_races$RACE = factor(metadata_all_races$RACE, levels=c("White", "Black or African American", "Asian", "American Indian or Alaska Native", "Unknown"))

# American Indian or Alaska Native and Unknown categories have very small sample sizes
metadata = metadata[! metadata$RACE %in% c("American Indian or Alaska Native", "Unknown"),]
metadata$RACE = factor(metadata$RACE, levels=c("White", "Black or African American", "Asian"))

# Change the factor levels to easily calculate coefficients for AA vs. Asian comparisons.
# Use AA for reference since there are more AA in the cohort
metadata_all_races_aa_ref = metadata_all_races
metadata_all_races_aa_ref$RACE = metadata_all_races_aa_ref$RACE = factor(metadata_all_races$RACE, levels=c("Black or African American", "Asian", "American Indian or Alaska Native", "Unknown", "White"))
```


## Functions
```{r}
postprocess_glm <- function(t5) {
  pdat = as.data.frame(t5)

  # Calculate 95% confidence intervals
  pdat$cil = pdat$Estimate + qnorm(0.025)*pdat$`Std. Error`
  pdat$ciu = pdat$Estimate + qnorm(0.975)*pdat$`Std. Error`
  
  # Create user-friendly factor level names
  tmp = strsplit(rownames(pdat), ":", fixed = TRUE)
  pdat$tissue = sapply(tmp, "[[", 1)
  pdat$var2 = sapply(tmp, "[[", 2)
  rownames(pdat) = NULL
  pdat$tissue = sub("^SMTSD", "", pdat$tissue)
  
  pdat$var2_type = pdat$var2
  pdat$var2_type[grepl("^RACE", pdat$var2)] = "Ancestry"
  pdat$var2_type[grepl("^SEX", pdat$var2)] = "SEX"
  
  pdat$var2 = sub("^RACE", "", pdat$var2)
  pdat$var2[pdat$var2 == "Black or African American"] = "African American"
  pdat$var2 = sub("^SEX", "", pdat$var2)
  
  pdat$type_pretty = pdat$var2
  
  # Separate by covariate type
  pdat.age = pdat[which(pdat$type_pretty=="AGE"),]
  pdat.sex = pdat[which(pdat$var2_type=="SEX"),]
  pdat.ancestry = pdat[which(pdat$var2_type=="Ancestry"),]
  
  return(list(pdat.age=pdat.age, pdat.sex=pdat.sex, pdat.ancestry=pdat.ancestry))
}

my_plot <- function(pdat.ancestry, i, mega_pdat.ancestry, reference_level, save_fig_bool, width, height, today, fig_dir, model_name) {
    # Get the coordinates if these were plotted on the same axes
    p = ggplot(data=mega_pdat.ancestry, aes(y=tissue, x=Estimate, xmin=cil, xmax=ciu)) +
      geom_errorbarh(height=0) + 
      geom_point(aes(fill=tissue), size=2.5, shape=21, col="black") +
      tissue_palette$fill_scale +
      facet_grid(type_pretty ~., scales="free", space="free") +
      geom_vline(xintercept=0,color="black",linetype="dashed")
    
    my_x_limits = layer_scales(p)$x$get_limits()
  
    tmp = pdat.ancestry[pdat.ancestry$type_pretty == i,]
    tmp = tmp[with(tmp, order(Estimate)),]
    tmp$tissue = factor(tmp$tissue, levels=tmp$tissue)
  
    if (save_fig_bool) {
      fn = sanitize_var_for_fn(file.path(fig_dir, sprintf("%s.tissue_ancestry_effects.%s_vs_%s.%s.pdf", model_name, i, reference_level, today)))
      pdf(fn, width=width, height=height)
    }
    
    p = ggplot(data=tmp, aes(y=tissue, x=Estimate, xmin=cil, xmax=ciu)) +
      geom_errorbarh(height=0) + 
      geom_point(aes(fill=tissue), size=2.5, shape=21, col="black") +
      tissue_palette$fill_scale +
      geom_vline(xintercept=0,color="black",linetype="dashed") +  
      xlab("Beta") +
      ylab("Tissue") +
      ggtitle(sprintf("%s vs. %s", i, reference_level)) +
      my_theme() +
      scale_x_continuous(limits=my_x_limits) +
      theme(panel.spacing = unit(1, "lines"),
            legend.position = "none",
            text=element_text(size=14),
            panel.grid=element_blank(),
            strip.background = element_blank(),
            plot.margin = margin(5.5, 30, 5., 5.5, unit="pt"))
  grid.newpage()
  p2 = timestamp_ggfigure(p)
  grid.draw(p2)
  
  if (save_fig_bool) {
    dev.off()
    write_new_file_message(fn)
  }

  grid.draw(p2)
  
  return()
}

plot_sex <- function(t5, is_transformed, save_fig_bool, width, height, today, fig_dir, model_name) {
  tmp = t5$pdat.sex
  tmp = tmp[with(tmp, order(Estimate)),]
  tmp$tissue = factor(tmp$tissue, levels=tmp$tissue)

  if (is_transformed) {
    suffix = "transformed_beta"
    x_lab = "Difference in # transcribed PZMs"
  } else {
    suffix = ""
    x_lab = "Beta"
  }
  
  if (save_fig_bool) {
    fn = sanitize_var_for_fn(file.path(fig_dir, sprintf("%s.tissue_sex_effects%s.%s.pdf", model_name, suffix, today)))
    pdf(fn, width=width, height=height)
  }
  
  p = ggplot(data=tmp, aes(y=tissue, x=Estimate, xmin=cil, xmax=ciu)) +
    geom_errorbarh(height=0) + 
    geom_point(aes(fill=tissue), size=2.5, shape=21, col="black") +
    tissue_palette$fill_scale +
    geom_vline(xintercept=0,color="black",linetype="dashed") +  
    xlab(x_lab) +
    ylab("Tissue") +
    my_theme() +
    theme(panel.spacing = unit(1, "lines"),
          legend.position = "none",
          text=element_text(size=14),
          panel.grid=element_blank(),
          strip.background = element_blank())
  grid.newpage()
  p2 = timestamp_ggfigure(p)
  grid.draw(p2)
  
  if (save_fig_bool) {
    dev.off()
    write_new_file_message(fn)
  }
  
  grid.draw(p2)
}

plot_age <- function(t5, save_fig_bool, width, height, today, fig_dir, model_name) {
  tmp = t5$pdat.age
  tmp = tmp[with(tmp, order(Estimate)),]
  tmp$tissue = factor(tmp$tissue, levels=tmp$tissue)
  
  if (save_fig_bool) {
    fn = sanitize_var_for_fn(file.path(fig_dir, sprintf("%s.tissue_age_effects.%s.pdf", model_name, today)))
    pdf(fn, width=width, height=height)
  }
  
  p = ggplot(data=tmp, aes(y=tissue, x=Estimate, xmin=cil, xmax=ciu)) +
    geom_errorbarh(height=0) + 
    geom_point(aes(fill=tissue), size=2.5, shape=21, col="black") +
    tissue_palette$fill_scale +
    geom_vline(xintercept=0,color="black",linetype="dashed") +  
    xlab("Beta") +
    ylab("Tissue") +
    my_theme() +
    theme(panel.spacing = unit(1, "lines"),
          legend.position = "none",
          text=element_text(size=14),
          panel.grid=element_blank(),
          strip.background = element_blank(),
          plot.margin = margin(5.5, 10, 5.5, 5.5, unit="pt"))
  grid.newpage()
  p2 = timestamp_ggfigure(p)
  grid.draw(p2)
  
  if (save_fig_bool) {
    dev.off()
    write_new_file_message(fn)
  }
  
  grid.draw(p2)
}

plot_f <- function(ares, save_fig_bool, width, height, today, fig_dir, model_name) {
  # Sort by decreasing F statusitic
  fdat = ares[order(ares[,3],decreasing=T),]
  
  # Remove residuals
  fdat = fdat[-nrow(fdat),]
  
  fdat = as.data.frame(fdat)
  fdat$variable = rownames(fdat)
  
  # Use read friendly variable names
  fdat$variable = revalue(fdat$variable, 
    c("avg_simulated_var_power"="Sample mutation power", 
      "AGE"="Age",
      "SMTSD"="Tissue",
      "num_mbp_w_ge20X"="Transcriptome size",
      "SMTSD:AGE"="Tissue:age",
      "SMRIN"="RIN",
      "SEX"="Sex",
      "SMNABTCH"="Batch",
      "genotype_data"="Genotype data",
      "SMTSD:RACE"="Tissue:ancestry",
      "SMTSD:SEX"="Tissue:sex",
      "RACE"="Ancestry"))
  
  fdat$variable = factor(fdat$variable, levels=fdat$variable)
  
  if (save_fig_bool) {
    fn = file.path(fig_dir, sprintf("%s.covariate_f_statistics.%s.pdf", model_name, today))
    pdf(fn, width=width, height=height)
  }
  p = ggplot(fdat, aes(x=variable, y=`F value`)) +
    geom_col(fill="black") +
    my_theme() +
    theme_classic() +
    theme(text=element_text(size=12),
          plot.margin = margin(5.5, 10, 5., 5.5, unit="pt")) + # x-axis was being cut off on right-hand side
    coord_flip() +
    xlab("Covariate") +
    ylab("F value")
  grid.newpage()
  p2 = timestamp_ggfigure(p)
  grid.draw(p2)
  
  if (save_fig_bool) {
    dev.off()
    write_new_file_message(fn)
  }
  
  grid.draw(p2)
}
```

# Model mutation burden
## Mutation burden = # of PZMs in sample
### Fit model
```{r}
model_name = "count_mutation_burden"

# These lm calls take a few minutes to run
fit3.1 = lm(log10(num_em_var + 1) ~ SMRIN + SMTSD + SMNABTCH + AGE + RACE + SEX + genotype_data + num_mbp_w_ge20X + avg_simulated_var_power + RACE:SMTSD + SEX:SMTSD + AGE:SMTSD, data=metadata_all_races)

fit3.1_aa = lm(log10(num_em_var + 1) ~ SMRIN + SMTSD + SMNABTCH + AGE + RACE + SEX + genotype_data + num_mbp_w_ge20X + avg_simulated_var_power + RACE:SMTSD + SEX:SMTSD + AGE:SMTSD, data=metadata_all_races_aa_ref)
```

```{r}
summary(fit3.1)
```


```{r}
# Get f values from fitted GLM
ares = Anova(fit3.1,type="II") # This takes ~15 minutes to run
```

```{r}
# Save 
output_file = file.path(output_dir, sprintf("model.%s.%s.rds", model_name, today))
saveRDS(fit3.1, file=output_file)
write_new_file_message(output_file)
```


### Plot
#### Plot F statistics
```{r, fig.height=3, fig.width=4}
plot_f(ares)
```


#### Plot tissue interaction betas
```{r}
t2 = summary(fit3.1)
# AA vs. asian comparison
t2_aa = summary(fit3.1_aa)

t3 = t2$coefficients
t4 = t3[grep(":",rownames(t3)),]
t5_unprocessed = t4[which(t4[,4] <= 0.05),]
row.names(t5_unprocessed)<-row.names(t4)[which(t4[,4] <= 0.05)]

t3_aa = t2_aa$coefficients
t4_aa = t3_aa[grep(":",rownames(t3_aa)),]
t5_aa_unprocessed = t4_aa[which(t4_aa[,4] <= 0.05),]
row.names(t5_aa_unprocessed)<-row.names(t4_aa)[which(t4_aa[,4] <= 0.05)]

t5_white = postprocess_glm(t5_unprocessed)

t5_aa = postprocess_glm(t5_aa_unprocessed)

mega_pdat.ancestry = rbind(
  t5_white$pdat.ancestry[! t5_white$pdat.ancestry$type_pretty %in% c("Unknown", "American Indian or Alaska Native"),],
  t5_aa$pdat.ancestry[t5_aa$pdat.ancestry$type_pretty == "Asian",])
```

##### Ancestry
```{r, fig.height=2, fig.width=5}
my_plot(t5_white$pdat.ancestry, i="African American", mega_pdat.ancestry, reference_level="White", save_fig_bool=save_fig_bool, width=5, height=2, today=today, fig_dir=fig_dir, model_name=model_name)
```

```{r, fig.height=3.5, fig.width=5}
p2 = my_plot(t5_white$pdat.ancestry, i="Asian", mega_pdat.ancestry, reference_level="White", save_fig_bool=save_fig_bool, width=5, height=3.5, today=today, fig_dir=fig_dir, model_name=model_name)
```

```{r, fig.height=2.75, fig.width=5}
p2 = my_plot(t5_aa$pdat.ancestry, i="Asian", mega_pdat.ancestry, reference_level="African American", save_fig_bool=save_fig_bool, width=5, height=2.75, today=today, fig_dir=fig_dir, model_name=model_name)
```

##### Sex
```{r, fig.height=3, fig.width=5.25}
plot_sex(t5_white, is_transformed=FALSE, save_fig_bool=save_fig_bool, width=5.25, height=3, today=today, fig_dir=fig_dir, model_name=model_name)
```

##### Age
```{r}
plot_age(t5_white, save_fig_bool=save_fig_bool, width=5, height=4, today=today, fig_dir=fig_dir, model_name=model_name)
```

### Estimate % of donor variance explained
```{r}
# Run repeatability estimation from generlized linear mixed-effects models fitted by restricted maximum likelihood using rptR's rpt function
# rpt was run in parallel on a cluster to decrease the amount of wall clock time.  See input_files/get_rpt_estimates.R.
rdat_fn = file.path(input_dir, "repeat_estimates.20210928.Rdata")
load(rdat_fn) # Loads rpt_combined_races, rpt_individual_races

# Grab data
tmp1 = do.call(rbind, lapply(names(rpt_combined_races), function(nboot) {
  data.frame(
    dataset = "All donors",
    nboot = nboot,
    estimate = rpt_combined_races[[nboot]]$R$SUBJID,
    ci_low = unname(rpt_combined_races[[nboot]]$CI_emp[1]),
    ci_high = unname(rpt_combined_races[[nboot]]$CI_emp[2])
  )
}))

tmp2 = do.call(rbind, lapply(names(rpt_individual_races), function(dataset) {
  do.call(rbind, lapply(names(rpt_individual_races[[dataset]]), function(nboot) {
    data.frame(
      dataset = dataset,
      nboot = nboot,
      estimate = rpt_individual_races[[dataset]][[nboot]]$R$SUBJID,
      ci_low = unname(rpt_individual_races[[dataset]][[nboot]]$CI_emp[1]),
      ci_high = unname(rpt_individual_races[[dataset]][[nboot]]$CI_emp[2])
    )
  }))
}))

donor_variance_df = rbind(tmp1, tmp2)

donor_variance_df$ci = donor_variance_df$ci_high - donor_variance_df$ci_low


# Add # of donors
tmp = unique(metadata[,c("SUBJID", "RACE")])
num_donors_df = as.data.frame(table(tmp$RACE))
colnames(num_donors_df) = c("dataset", "N")
num_donors_df = rbind(num_donors_df, data.frame(dataset="All donors", N=nrow(tmp)))

donor_variance_df = merge(donor_variance_df, num_donors_df, by="dataset")

donor_variance_df$dataset = factor(donor_variance_df$dataset, levels=
  c("All donors",
    "American Indian or Alaska Native",
    "Asian",
    "Black or African American",
    "White",
    "Unknown"))

donor_variance_df$dataset = revalue(donor_variance_df$dataset, c("Black or African American"="Black or\nAfrican\nAmerican"))

donor_variance_df = donor_variance_df[with(donor_variance_df, order(dataset)),]


donor_variance_df$dataset_pretty = sprintf("%s\nN = %d", donor_variance_df$dataset, donor_variance_df$N)
donor_variance_df$dataset_pretty = factor(donor_variance_df$dataset_pretty, levels=unique(donor_variance_df$dataset_pretty))
```

#### Plot
```{r, fig.height=3, fig.width=3}
if (save_fig_bool) {
  fn = file.path(fig_dir, sprintf("figure_1.donor_variance_explained.%s.pdf", today))
  pdf(fn, width=3, height=3)
}
scaleFUN <- function(x) sprintf("%.0f%%", x*100)

tmp = donor_variance_df[donor_variance_df$nboot == "500" &
      ! donor_variance_df$dataset %in% c("Asian", "Unknown"),]
  
p = ggplot(tmp, aes(x=dataset_pretty, y=estimate))+
  geom_point(size=3) + 
  geom_errorbar(width=0, aes(ymin=ci_low, ymax=ci_high)) +
  scale_y_continuous(labels=scaleFUN) +
  xlab("Self-reported ancestry") +
  ylab("% of variance explained\nby donor effects") +
  my_theme() +
  theme_classic() +
  theme(text=element_text(size=12))+
  expand_limits(y=0)
grid.newpage()
p2 = timestamp_ggfigure(p)
grid.draw(p2)

if (save_fig_bool) {
  dev.off()
  write_new_file_message(fn)
}

grid.draw(p2)
```

## Mutation burden = VAF sum
### Fit model
```{r}
model_name = "vaf_sum_mutation_burden"

# These lm calls take a few minutes to run
fit4 = lm(log10(sample_vaf_sum_pseudo) ~ SMRIN + SMTSD + SMNABTCH + AGE + RACE + SEX + genotype_data + num_mbp_w_ge20X + avg_simulated_var_power + RACE:SMTSD + SEX:SMTSD + AGE:SMTSD, data=metadata_all_races)

# Change the factor levels to easily calculate coefficients for AA vs. Asian comparisons.
fit4_aa = lm(log10(sample_vaf_sum_pseudo) ~ SMRIN + SMTSD + SMNABTCH + AGE + RACE + SEX + genotype_data + num_mbp_w_ge20X + avg_simulated_var_power + RACE:SMTSD + SEX:SMTSD + AGE:SMTSD, data=metadata_all_races_aa_ref)
```

```{r}
summary(fit4)
```

```{r}
# Get f values from fitted GLM
ares4 = Anova(fit4,type="II") # This takes ~15 minutes to run
```

```{r}
# Save 
output_file = file.path(output_dir, sprintf("model.%s.%s.rds", model_name, today))
saveRDS(fit4, file=output_file)
write_new_file_message(output_file)
```

### Plot
#### Plot F statistics
```{r, fig.height=3, fig.width=4}
plot_f(ares4)
```

#### Plot tissue interaction betas
```{r}
t2 = summary(fit4)
# AA vs. asian comparison
t2_aa = summary(fit4_aa)

t3 = t2$coefficients
t4 = t3[grep(":",rownames(t3)),]
t5_unprocessed = t4[which(t4[,4] <= 0.05),]
row.names(t5_unprocessed)<-row.names(t4)[which(t4[,4] <= 0.05)]

t3_aa = t2_aa$coefficients
t4_aa = t3_aa[grep(":",rownames(t3_aa)),]
t5_aa_unprocessed = t4_aa[which(t4_aa[,4] <= 0.05),]
row.names(t5_aa_unprocessed)<-row.names(t4_aa)[which(t4_aa[,4] <= 0.05)]

t5_white = postprocess_glm(t5_unprocessed)

t5_aa = postprocess_glm(t5_aa_unprocessed)

mega_pdat.ancestry = rbind(
  t5_white$pdat.ancestry[! t5_white$pdat.ancestry$type_pretty %in% c("Unknown", "American Indian or Alaska Native"),],
  t5_aa$pdat.ancestry[t5_aa$pdat.ancestry$type_pretty == "Asian",])
```

##### Ancestry
```{r, fig.height=2, fig.width=5}
my_plot(t5_white$pdat.ancestry, i="African American", mega_pdat.ancestry, reference_level="White", save_fig_bool=save_fig_bool, width=5, height=2, today=today, fig_dir=fig_dir, model_name=model_name)
```

```{r, fig.height=3.5, fig.width=5}
p2 = my_plot(t5_white$pdat.ancestry, i="Asian", mega_pdat.ancestry, reference_level="White", save_fig_bool=save_fig_bool, width=5, height=3.5, today=today, fig_dir=fig_dir, model_name=model_name)
```

```{r, fig.height=2.75, fig.width=5}
p2 = my_plot(t5_aa$pdat.ancestry, i="Asian", mega_pdat.ancestry, reference_level="African American", save_fig_bool=save_fig_bool, width=5, height=2.75, today=today, fig_dir=fig_dir, model_name=model_name)
```

##### Sex
```{r, fig.height=3, fig.width=5.25}
plot_sex(t5_white, is_transformed=FALSE, save_fig_bool=save_fig_bool, width=5.25, height=3, today=today, fig_dir=fig_dir, model_name=model_name)
```

##### Age
```{r}
plot_age(t5_white, save_fig_bool=save_fig_bool, width=5, height=4, today=today, fig_dir=fig_dir, model_name=model_name)
```
