---
title: "Make figures related to comparing full tissue and germlayer trees"
output: html_notebook
---

# Metadata
* **Last modified:** `r format(Sys.time(), "%Y%m%d")`

## Purpose
* Plot full tissue and germlayer mutation buden trees 
    + w/ their own burden color legend to improve readability of within tree comparisons
    + w/ same burden color legend to improve readability of between tree comparisons
    + using "sum of descendent edges" mutation burden trees to improve readability of between tree comparisons
* Create table of edge weights

# Setup environment
```{r}
source("../tree_utils.R")
source("../utils.R")
source("mk_tree_mutation_burden.functions.R")
```

## Globals
```{r}
local = 2
today = get_todays_date()
input_dir = "input_files"
output_dir = "output_files"
na_color = "gray"
# save_file = FALSE
save_file = TRUE

tree_raw_weights = list()
tree_consensus_g = list()
```


# Full tree
```{r}
tree_name = "full.no_gtex_blacklist.no_sex.no_cell_lines"
consensus_g_fn = file.path(input_dir, "consensus_g.power_adjusted_weights.exact_match.full.no_gtex_blacklist.no_sex.no_cell_lines.20200714.rds")
edge_indexes_to_exclude = c(9, 46, 58)

tmp = load_data(tree_name, consensus_g_fn, local, edge_indexes_to_exclude)

tissue_g = tmp$tissue_g
consensus_g = tmp$consensus_g
tree_consensus_g[[tree_name]] = consensus_g
leaves = tmp$leaves
non_leaf_edge_indexes = tmp$non_leaf_edge_indexes
distance_from_root = tmp$distance_from_root
mutation_vector_str2edge = tmp$mutation_vector_str2edge
```

## Raw weights
```{r, fig.width=20, fig.height=20}
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.raw_weights.%s.%s.pdf", tree_name, today))

if (save_file) {
    pdf(fn, width=20, height=20)
}
consensus_g_raw_weights = plot_g(consensus_g, type="mutation_prob", category="numeric", save_file=FALSE, set_na_colors = TRUE, set_widths=FALSE, diverging_palette=FALSE, set_edge_labels=FALSE, na_color="grey", edge_label_color="black", set_vertex_labels=FALSE, use_log_trans = FALSE, pseudo_count = 0)
tree_raw_weights[[tree_name]] = consensus_g_raw_weights

if (save_file) {
    dev.off()
    write_new_file_message(fn)
}
```

Make a color bar
```{r, fig.height=7, fig.width=2}
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.raw_weights.%s.color_bar.fixed_color_bar_range.%s.pdf", tree_name, today))
mk_color_bar(consensus_g_raw_weights, fn, log_trans=FALSE, percent=TRUE, save_file=save_file)
```

## Log-scaled weights
```{r, fig.width=20, fig.height=20}
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.log_weights.%s.%s.pdf", tree_name, today))

if (save_file) {
    pdf(fn, width=20, height=20)
}
consensus_g_log_weights = plot_g(consensus_g, type="mutation_prob", category="numeric", save_file=FALSE, set_na_colors = TRUE, set_widths=FALSE, diverging_palette=FALSE, set_edge_labels=FALSE, na_color="grey", edge_label_color="black", set_vertex_labels=FALSE, use_log_trans = TRUE, pseudo_count = 0)
full_tree_consensus_g_log_weights = consensus_g_log_weights
if (save_file) {
    dev.off()
    write_new_file_message(fn)
}
```

Make a color bar
```{r, fig.height=7, fig.width=2}
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.log.weights.%s.color_bar.%s.pdf", tree_name, today))
mk_color_bar(consensus_g_log_weights, fn, log_trans=TRUE, percent=TRUE, save_file=save_file)
```

## Create a table of edge weights
```{r}
unknown_weights = E(consensus_g)$weight_unknown
edge_mutation_weights = E(consensus_g)$mutation_prob

edge_weight_df = as.data.frame(ends(consensus_g, E(consensus_g), names=TRUE))
colnames(edge_weight_df) = c("edge_head", "edge_tail")
edge_weight_df$edge_mutation_weight = edge_mutation_weights
edge_weight_df$edge_mutation_weight[unknown_weights] = NA
```

```{r}
output_fn = file.path(output_dir, sprintf("mutation_burden.weights.%s.%s.txt", tree_name, today))
write.table(edge_weight_df, file=output_fn, row.names=FALSE, quote=FALSE, sep="\t")
write_new_file_message(output_fn)
```

# Germlayer tree
```{r}
tree_name = "germ_layers"
consensus_g_fn = file.path(input_dir, "consensus_g.power_adjusted_weights.exact_match.germ_layers.20200715.rds")
edge_indexes_to_exclude = c()

tmp = load_data(tree_name, consensus_g_fn, local, edge_indexes_to_exclude)

tissue_g = tmp$tissue_g
consensus_g = tmp$consensus_g
tree_consensus_g[[tree_name]] = consensus_g
leaves = tmp$leaves
non_leaf_edge_indexes = tmp$non_leaf_edge_indexes
distance_from_root = tmp$distance_from_root
mutation_vector_str2edge = tmp$mutation_vector_str2edge
```

## Raw weights
```{r, fig.width=20, fig.height=20}
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.raw_weights.%s.%s.pdf", tree_name, today))

if (save_file) {
    pdf(fn, width=20, height=20)
}
consensus_g_raw_weights = plot_g(consensus_g, type="mutation_prob", category="numeric", save_file=FALSE, set_na_colors = TRUE, set_widths=FALSE, diverging_palette=FALSE, set_edge_labels=FALSE, na_color="grey", edge_label_color="black", set_vertex_labels=FALSE, use_log_trans = FALSE, pseudo_count = 0)
tree_raw_weights[[tree_name]] = consensus_g_raw_weights

if (save_file) {
    dev.off()
    write_new_file_message(fn)
}
```


Make a color bar
```{r, fig.height=7, fig.width=2}
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.raw_weights.%s.color_bar.%s.pdf", tree_name, today))
mk_color_bar(consensus_g_raw_weights, fn, log_trans=FALSE, percent=TRUE, save_file=save_file)
```

## Create a table of edge weights
```{r}
unknown_weights = E(consensus_g)$weight_unknown
edge_mutation_weights = E(consensus_g)$mutation_prob

edge_weight_df = as.data.frame(ends(consensus_g, E(consensus_g), names=TRUE))
colnames(edge_weight_df) = c("edge_head", "edge_tail")
edge_weight_df$edge_mutation_weight = edge_mutation_weights
edge_weight_df$edge_mutation_weight[unknown_weights] = NA
```

```{r}
output_fn = file.path(output_dir, sprintf("mutation_burden.weights.%s.%s.txt", tree_name, today))
write.table(edge_weight_df, file=output_fn, row.names=FALSE, quote=FALSE, sep="\t")
write_new_file_message(output_fn)
```

# Use same color bar range for both trees
```{r}
# Calculate global min and max edge weight
min_weight = min(
    min(E(tree_raw_weights[["full.no_gtex_blacklist.no_sex.no_cell_lines"]])$values_used, na.rm=TRUE), 
    min(E(tree_raw_weights[["germ_layers"]])$values_used, na.rm=TRUE)) # 0 (0 is used for min in plot_g when values are >= 0)

# To set the max appropariate, calculate the raw weights for both trees then take the max
max_weight = max(
    max(E(tree_raw_weights[["full.no_gtex_blacklist.no_sex.no_cell_lines"]])$values_used, na.rm=TRUE), 
    max(E(tree_raw_weights[["germ_layers"]])$values_used, na.rm=TRUE)) # 0.44
```


```{r, fig.width=20, fig.height=20}
for (tree_name in names(tree_raw_weights)) {
    fn = file.path(output_dir, sprintf("mutation_burden.no_labels.raw_weights.%s.fixed_color_bar_range.%s.pdf", tree_name, today))
    
    consensus_g = tree_raw_weights[[tree_name]]
    
    if (save_file) {
        pdf(fn, width=20, height=20)
    }
    consensus_g_raw_weights = plot_g(consensus_g, type="mutation_prob", category="numeric", save_file=FALSE, set_na_colors = TRUE, set_widths=FALSE, diverging_palette=FALSE, set_edge_labels=FALSE, na_color="grey", edge_label_color="black", set_vertex_labels=FALSE, use_log_trans = FALSE, pseudo_count = 0, value_global_max=max_weight)
    
    if (save_file) {
        dev.off()
        write_new_file_message(fn)
    }
}
```

Make a color bar
```{r, fig.height=7, fig.width=2}
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.raw_weights.fixed_color_bar_range.color_bar.%s.pdf", today))
# Same color bar used for both trees so using last value of consensus_g_raw_weights is fine
mk_color_bar(consensus_g_raw_weights, fn, log_trans=FALSE, percent=TRUE, save_file=save_file, use_custom_range = TRUE, my_min=min_weight, my_max=max_weight)
```


# Create aggregate burden trees
## Calculate aggregate burden
```{r}
for (tree_name in names(tree_consensus_g)) {
    consensus_g = tree_consensus_g[[tree_name]]
    consensus_g = set.edge.attribute(consensus_g, "descendent_weight_sum", value=0)
    for (my_edge in E(consensus_g)) {
        descendent_nodes = get_descendants(consensus_g, head_of(consensus_g, my_edge)) 
        
        descendent_edges = c(my_edge, E(consensus_g)[from(descendent_nodes)])
        descendent_weight_sum = sum(get.edge.attribute(consensus_g, "mutation_prob", descendent_edges))
        consensus_g = set.edge.attribute(consensus_g, "descendent_weight_sum", my_edge, descendent_weight_sum)
    }
    tree_consensus_g[[tree_name]] = consensus_g
}
```

## Plot
```{r, fig.width=20, fig.height=20}
for (tree_name in names(tree_consensus_g)) {
    fn = file.path(output_dir, sprintf("mutation_burden.no_labels.descendent_weight_sum.%s.%s.pdf", tree_name, today))
    
    if (save_file) {
        pdf(fn, width=20, height=20)
    }
    consensus_g_sum_weights = plot_g(tree_consensus_g[[tree_name]], type="descendent_weight_sum", category="numeric", save_file=FALSE, set_na_colors = TRUE, set_widths=FALSE, diverging_palette=FALSE, set_edge_labels=FALSE, na_color="grey", edge_label_color="black", set_vertex_labels=FALSE, use_log_trans = FALSE, pseudo_count = 0)
    
    if (save_file) {
        dev.off()
        write_new_file_message(fn)
    }
}
```

Make a color bar
```{r, fig.height=7, fig.width=2}
# Same color bar used for both trees (both have, by definition, range [0,1] so using last value of consensus_g_raw_weights is fine
fn = file.path(output_dir, sprintf("mutation_burden.no_labels.descendent_weight_sum.color_bar.%s.pdf", tree_name, today))
mk_color_bar(consensus_g_sum_weights, fn, log_trans=FALSE, percent=TRUE, save_file=save_file)
```



